\documentclass{article}

\usepackage[nonatbib]{neurips}
\usepackage[numbers]{natbib}

\input{extra_pkgs}

\usepackage{physics}
\usepackage{mathtools}
\usepackage{multirow}
\usepackage{subcaption}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{xspace}

\usepackage{booktabs}
\usepackage{multirow}
% \usepackage{rotating}   % for sidewaystable*
\usepackage{graphicx}   % for \resizebox
\usepackage[table]{xcolor}

\newcommand{\method}{REIN\xspace}
\newcommand{\eg}{e.g.,\xspace}
\newcommand{\ie}{i.e.,\xspace}
\newcommand{\etal}{\textit{et al.}\xspace}

\title{REIN: Residual Modulation of Time Series Forecasts\\ via LLM-Refined News Regime Signals}

\author{
    Anonymous Author(s)
}

\begin{document}

\maketitle

%% ============================================================
%% ABSTRACT
%% ============================================================
\begin{abstract}
Time series forecasting remains central to modern decision-making, and news text is increasingly valuable for adjusting such forecasts. We introduce \method (\textbf{RE}sidual \textbf{I}nformation from \textbf{N}ews), a two-stage framework that augments time series forecasting with LLM-refined textual signals through multiplicative residual modulation.
Given a corpus of filtered news articles, \method first distills each article into a structured regime vector via a large language model, then aggregates these vectors into a daily regime bank.
A PatchTST-based delta module decomposes the base model's residual into three components capturing distinct temporal patterns, whose magnitudes are modulated by the news regime representation. When no relevant news exists, the modulation knobs collapse to identity and the system gracefully degrades to the base forecaster with bounded perturbation.
To prevent spurious reliance on news during training, we adopt a counterfactual protocol with news-blanking and date-permutation tests, verifying that gains are causally attributable to news content rather than calendar artifacts.
We evaluate \method on five diverse datasets such as Australian electricity load, Australian road traffic, European natural gas demand, Bitcoin hourly prices, and Australian electricity price. Each paired with domain news curated from the GDELT global news corpus. Across horizons from 48 to 720 steps, \method achieves state-of-the-art performance, outperforming time-series, foundation, and LLM-based baselines.
\end{abstract}

%% ============================================================
%% 1. INTRODUCTION
%% ============================================================
\section{Introduction}
\label{sec:intro}

Accurate time series forecasting underpins decision-making in energy markets, transportation, and finance.
In modern society, where information propagates almost instantaneously through global news networks, numerical histories alone rarely capture the full picture of what drives a series. A policy announcement, an unexpected outage, a geopolitical event, or a viral social signal can reshape demand, traffic, or price within hours, long before such shifts are reflected in purely statistical patterns. Human analysts routinely consult news feeds to contextualize their forecasts, adjusting expectations for regulatory rulings, supply disruptions, or shifts in market sentiment. As news generation accelerates and real-world datasets become increasingly multimodal, incorporating textual signals into time series forecasting has shifted from a peripheral enhancement to a practical necessity for reliable and timely predictions.

While modern deep forecasters~\citep{PLACEHOLDER_patchtst_nie2023, PLACEHOLDER_dlinear_zeng2023, PLACEHOLDER_informer_zhou2021} achieve strong performance by exploiting temporal patterns, they remain blind to exogenous textual information such as weather warnings, policy announcements, and supply disruptions that human analysts routinely incorporate.
Integrating such unstructured text into numerical forecasters is challenging for three reasons:
(1)~raw news is noisy, with most articles irrelevant to the target series;
(2)~the causal link from text to time series varies across domains and time scales;
(3)~naive integration risks \emph{news hallucination}, where the model fabricates spurious correlations that degrade forecasts when news is absent.

Recent work has explored incorporating textual information into time series models through prompt-based LLM forecasters~\citep{PLACEHOLDER_llmtime_gruver2024, PLACEHOLDER_timellm_jin2024}, cross-modal fusion~\citep{PLACEHOLDER_crossformer_zhang2023}, and retrieval-augmented approaches~\citep{PLACEHOLDER_rag_ts_2024}.
However, these approaches typically treat text as an additive input feature, offering no formal control over its influence and no guarantee of graceful degradation when relevant news is missing.

We propose \method, a two-stage framework that augments time series forecasting with refined textual signals through a principled gating mechanism that promotes useful news injection while suppressing irrelevant noise. In the first stage, a large language model distills each raw article into a structured regime vector, and these vectors are temporally aggregated into a daily regime bank that serves as a compact, denoised representation of contemporaneous news. In the second stage, a PatchTST-based delta module decomposes the base forecaster's residual into three components capturing distinct temporal patterns (trend correction, shape deformation, and spike behavior), and the news regime modulates the magnitude of each component multiplicatively rather than additively. Under this design, when no relevant news is present, the modulation factors collapse to identity values, and the system reduces to the base forecaster with provably bounded perturbation.

Our contributions are summarized as follows.
\begin{itemize}[leftmargin=1.5em, itemsep=2pt]
    \item \textbf{Structured news distillation via LLM refinement.}
    We design an LLM-based refinement pipeline that transforms noisy raw articles into fixed-schema regime vectors consisting of five continuous intensity dimensions, fifteen topic tags, and a 384-dimensional semantic embedding. The structured representation filters out irrelevant content and preserves actionable signals in a form directly consumable by downstream forecasters.

    \item \textbf{Multiplicative residual modulation with bounded degradation.}
    Rather than injecting news additively, \method constrains news to act only as a multiplicative modulator on residual components. Three bounded knobs are introduced: $\lambda_{\text{base}}$ for base-model trust, $g_{\text{shape}}$ for shape gain, and $b_{\text{spike}}$ for spike bias. When no news is active, all knobs collapse to identity values ($\lambda \to \lambda_{\min}$, $g \to 1$, $b \to 0$), yielding a formal degradation bound that provably prevents news hallucination.

    \item \textbf{Counterfactual evaluation protocol for causal attribution.}
    We introduce a three-condition protocol that evaluates each test sample under its normal input, an all-zero news-blanked input, and a date-permuted input with shuffled news-to-date alignment. This protocol isolates whether observed improvements stem from genuine news content or from confounding calendar effects, providing causal rather than merely correlational evidence.

    \item \textbf{Empirical validation across five heterogeneous domains.}
    We benchmark \method on five datasets covering electricity load, road traffic, natural gas demand, cryptocurrency prices, and electricity prices, each paired with a domain-appropriate news corpus. Across forecasting horizons from 48 to 720 steps, \method achieves state-of-the-art performance and outperforms time-series, foundation-model, and LLM-based baselines.
\end{itemize}

We evaluate \method on five datasets: Australian electricity load (NSW), Australian road traffic counts (NSW, permanent stations), European natural gas demand (Netherlands), Bitcoin hourly prices, and Australian electricity price (NSW).
Each dataset is paired with a domain-appropriate news corpus, including curated energy and renewables outlets for load and price, CBS archival news for gas demand, and the open GDELT global news corpus~\citep{PLACEHOLDER_gdelt_leetaru2013} for traffic and Bitcoin, with forecast horizons extending up to 720 steps.
Across the four datasets already re-trained under the leakage-free regime bank, overall MAE improves over a frozen DLinear base by 9.2\% (NSW Load), 21.4\% (NSW Traffic), 5.1\% (NSW Price), and 2.3\% (NL Gas), with top-decile help rates in the 78--93\% range, indicating that the delta module is most active precisely where the base forecaster struggles.
The counterfactual protocol confirms causal attribution in every case: blanking news hurts active samples by 2--26\% while the inactive blank gap stays within 3\% of the base forecast.
Bitcoin is still being re-trained under a newly built hourly regime bank (a sub-daily leakage path existed in the previous day-level bank) and its numbers are intentionally omitted in this revision.



%% ============================================================
%% 2. RELATED WORK
%% ============================================================
\section{Related Work}
\label{sec:related}

\paragraph{Deep time series forecasting.}
Transformer-based models have become dominant in long-horizon forecasting.
Informer~\citep{PLACEHOLDER_informer_zhou2021} and Autoformer~\citep{PLACEHOLDER_autoformer_wu2021} first adapted sparse and decomposition-aware attention to long sequences, while FEDformer~\citep{PLACEHOLDER_fedformer_zhou2022} moved the attention operation into the frequency domain to better capture periodic structure.
PatchTST~\citep{PLACEHOLDER_patchtst_nie2023} subsequently introduced channel-independent patching with self-attention, achieving state-of-the-art results on standard benchmarks by treating each variate as an independent token sequence.
In parallel, DLinear~\citep{PLACEHOLDER_dlinear_zeng2023} demonstrated that simple linear models with trend-seasonal decomposition can match or surpass Transformer performance, sparking renewed interest in lightweight architectures.
iTransformer~\citep{PLACEHOLDER_itransformer_liu2024} further inverted the attention dimension so that tokens correspond to variates rather than time steps, enabling explicit modeling of cross-variate correlations.
More recently, time series foundation models such as TimesFM~\citep{PLACEHOLDER_timesfm_das2024} and Chronos~\citep{PLACEHOLDER_chronos_ansari2024} pursue zero-shot transfer via large-scale pretraining on heterogeneous corpora.
All of these methods operate purely on numerical histories, leaving no mechanism to ingest exogenous textual context.
\method is complementary to this line of work: it uses an MLP or DLinear base backbone for stable temporal extrapolation, while a PatchTST-style delta module is tasked with correcting residuals conditioned on news, cleanly separating what the base can learn from what news can correct.

\paragraph{Text-augmented forecasting.}
Incorporating textual signals for time series prediction has a long history in finance~\citep{PLACEHOLDER_sentiment_finance_2020} and energy~\citep{PLACEHOLDER_news_energy_2022}, where early work relied on sentiment scores, topic models, or hand-crafted event indicators as exogenous features.
With the rise of pretrained language models, a new wave of methods directly couples time series with LLMs.
LLMTime~\citep{PLACEHOLDER_llmtime_gruver2024} repurposes general-purpose LLMs as zero-shot forecasters by tokenizing numerical values, while Time-LLM~\citep{PLACEHOLDER_timellm_jin2024} aligns time series patches with language model representations via reprogramming layers.
GPT4TS~\citep{PLACEHOLDER_gpt4ts_zhou2023} freezes the LLM backbone and fine-tunes only lightweight adapters for forecasting, imputation, and classification.
UniTime~\citep{PLACEHOLDER_unitime_liu2024} unifies multiple forecasting domains under a single model with language-guided instructions, and cross-modal fusion approaches~\citep{PLACEHOLDER_crossformer_zhang2023} interleave text embeddings with numerical features inside a shared encoder.
Retrieval-augmented variants~\citep{PLACEHOLDER_rag_ts_2024} further condition predictions on retrieved articles at inference time.
Despite this diversity, existing methods share two limitations: they either require an LLM in the inference loop, which is prohibitive for high-frequency or long-horizon deployment, or they inject textual features additively without any formal control over their influence, leaving the forecaster vulnerable to noisy or irrelevant news.
\method departs from these approaches in two key respects. It uses the LLM \emph{only offline} as a news distiller, yielding a lightweight inference pipeline, and it enforces a strictly multiplicative coupling between news and residual components, which guarantees graceful degradation to the base forecaster whenever relevant news is absent.



%% ============================================================
%% 3. METHOD
%% ============================================================
\section{Method}
\label{sec:method}

\subsection{Overview}
\label{sec:method_overview}

Given a univariate time series $\{y_t\}$ and an associated news corpus $\mathcal{N}$, \method produces a forecast $\hat{\bm{y}}$ through an additive decomposition into a \emph{base} prediction and a news-modulated \emph{delta} correction.
The \textbf{base forecaster} is a standard deep time series model (PatchTST, DLinear, NLinear, or MLP) trained purely on numerical histories. It captures the bulk of temporal regularity, such as trend, seasonality, and calendar effects, and is frozen after Stage~1, providing a stable reference prediction $\hat{\bm{y}}^{\text{base}}$ together with an internal feature $\bm{f}_{\text{base}}$.
The \textbf{delta module} is a lightweight PatchTST-based corrector trained in Stage~2. It operates on the base model's residual and predicts a structured correction term whose magnitude is multiplicatively modulated by news signals derived offline from $\mathcal{N}$. The final prediction in normalized (z-)space can be written as
\begin{equation}
    \hat{\bm{y}}_z \;=\; \hat{\bm{y}}^{\text{base}}_z \;+\; \Delta\!\big(\bm{h}_z,\, \bm{f}_{\text{base}},\, \mathcal{B}(d)\big),
    \label{eq:overview}
\end{equation}
where $\bm{h}_z$ is the history window, $d$ is the forecast origin date, and $\mathcal{B}$ is a daily regime bank constructed from the refined news corpus.

The central design constraint is that $\mathcal{B}(d)$ may only \emph{scale} the delta module's outputs and never inject additive news predictions. When no relevant news exists on day $d$, all modulation factors collapse to identity values and $\Delta$ reduces to $\lambda_{\min} \cdot \Delta_{\text{TS}}$, where $\Delta_{\text{TS}}$ is the purely time-series-derived correction and $\lambda_{\min}$ is a small constant. This guarantees graceful degradation: absent informative news, \method cannot deviate meaningfully from the base forecaster.

The remainder of this section details the base backbone (Sec.~\ref{sec:base}), the offline news distillation pipeline (Sec.~\ref{sec:news}), the delta module and its multiplicative modulation (Sec.~\ref{sec:delta}), the training procedure (Sec.~\ref{sec:training}), and the counterfactual protocol used to verify that observed improvements are causally attributable to news content (Sec.~\ref{sec:counterfactual}).

\subsection{Stage 1: Base Backbone}
\label{sec:base}

The base model maps a normalized history window $\bm{h}_z \in \mathbb{R}^{L}$ to a prediction $\hat{\bm{y}}^{\text{base}}_z \in \mathbb{R}^{H}$ and an internal representation $\bm{f}_{\text{base}} \in \mathbb{R}^{D_{\text{base}}}$, where $L$ is the look-back length and $H$ is the forecast horizon.
We apply robust quantile normalization $h_z = (h - \tilde{\mu}) / \tilde{\sigma}$, where $\tilde{\mu}$ is the training-set median and $\tilde{\sigma} = \text{IQR} / 1.349$, which provides robustness to extreme values such as price spikes.
The base is trained with Smooth-L1 loss in z-space (or a winsorized quantile loss for price) and is subsequently \emph{frozen} for all downstream usage, so that the delta module cannot compensate for weaknesses by silently reshaping the base prediction.
The backbone is chosen per dataset from a pool of PatchTST~\citep{PLACEHOLDER_patchtst_nie2023}, DLinear~\citep{PLACEHOLDER_dlinear_zeng2023}, NLinear, and MLP variants (cf.\ Section~\ref{sec:setup}).


\subsection{News Distillation Pipeline}
\label{sec:news}

Raw news articles are transformed into a structured representation through two offline stages.

\paragraph{Schema refinement.}
Each article is processed by an LLM with a structured prompt that extracts:
(i)~an actionability flag (non-actionable articles are permanently discarded);
(ii)~a 5-dimensional regime vector $\bm{r} \in [-1,1]^5$ covering market tightness, demand outlook, renewable surplus, volatility tone, and policy status;
(iii)~multi-label topic tags from a closed vocabulary of 15 categories (\eg \texttt{heatwave}, \texttt{supply\_tight}, \texttt{outage});
(iv)~an estimated influence horizon in days;
(v)~a self-assessed confidence score $c \in [0,1]$.
This schema is domain-specific but fixed once designed, requiring no per-article human annotation.

\paragraph{Regime bank construction.}
The refined articles are aggregated into a daily regime bank $\mathcal{B}$.
For each calendar day $d$, all articles with $\text{published\_at} \leq d \leq \text{published\_at} + \text{horizon\_days}$ and $\text{is\_actionable} = \text{true}$ are collected.
Each article receives a time-decayed weight $w = c \cdot \exp(-\text{age} / \tau)$, where $\tau = 5$ days.
The weighted average produces a daily regime vector $\bar{\bm{r}}_d$, topic mass $\bar{\bm{t}}_d \in \mathbb{R}^{15}$, and text embedding $\bar{\bm{e}}_d \in \mathbb{R}^{384}$ (from E5-small-v2~\citep{PLACEHOLDER_e5_wang2024}).
Temporal smoothing via exponential moving average ($\alpha = 0.5$) produces the final bank entries.
A \emph{relevance mass} $m_d = \sum_i w_i$ (unnormalized) quantifies the aggregate signal strength for day $d$; days with $m_d \leq \theta$ are treated as news-inactive.


\subsection{Stage 2: Delta Module}
\label{sec:delta}

Given the frozen base prediction $\hat{\bm{y}}^{\text{base}}_z$ and its internal feature $\bm{f}_{\text{base}}$, the delta module predicts a structured residual that corrects the base forecast.
It consists of four components: a time-series encoder, a regime projector, three residual heads, and three modulation heads. News only enters through the modulation heads, preserving the separation between what the base can learn from numerical history and what news can correct.

\paragraph{Time-series encoder.}
We employ a PatchTST-style encoder~\citep{PLACEHOLDER_patchtst_nie2023}.
The history $\bm{h}_z$ is split into overlapping patches of length $P{=}8$ with stride $S{=}4$, yielding $N_p = 1 + \lfloor(L-P)/S\rfloor$ patches.
Each patch is linearly projected to $D{=}128$ dimensions and augmented with learnable calendar embeddings (day-of-week, half-hour-of-day, holiday).
The frozen base representation $\bm{f}_{\text{base}}$ is prepended as an additional token so that the delta encoder can condition on what the base already captures.
A 2-layer Transformer encoder with 4 heads processes the sequence, producing per-token features $\bm{T} \in \mathbb{R}^{(N_p+1) \times D}$ and a mean-pooled summary $\bm{s} \in \mathbb{R}^{D}$.

\paragraph{Regime projector.}
The concatenation of $\bar{\bm{r}}$, $\bar{\bm{t}}$, and $\bar{\bm{e}}$ (total 404 dimensions) is projected through a two-layer MLP with GELU activations and LayerNorm to produce $\bm{v}_{\text{regime}} \in \mathbb{R}^{D}$.
An active mask $a = \mathbf{1}[m > \theta]$ gates this representation: when news is inactive, $\bm{v}_{\text{regime}} = \bm{0}$.

\paragraph{Three-head residual decomposition.}
The time-series features $\bm{T}$ and $\bm{s}$ are processed by three specialized heads:
\begin{align}
    \text{Slow head:} \quad & s_{\text{slow}} = f_{\text{slow}}(\bm{s}) \in \mathbb{R}, \label{eq:slow} \\
    \text{Shape head:} \quad & \bm{q}_{\text{shape}} = f_{\text{shape}}(\bm{T}) - \overline{f_{\text{shape}}(\bm{T})} \in \mathbb{R}^{H}, \label{eq:shape} \\
    \text{Spike head:} \quad & \bm{q}_{\text{spike}}, \bm{g}_{\text{logit}} = f_{\text{spike}}(\bm{T}) \in \mathbb{R}^{H} \times \mathbb{R}^{H}, \label{eq:spike}
\end{align}
where the slow head captures level shifts (broadcast across the horizon), the shape head outputs a zero-mean intra-day pattern, and the spike head produces both spike amplitudes and gating logits.
Training targets for these heads are derived from the true residual $\bm{y}_z - \hat{\bm{y}}^{\text{base}}_z$ by decomposition against a pre-computed calendar baseline ($7 \times 48$ day-of-week $\times$ half-hour lookup table) and quantile-based spike thresholds.

\paragraph{Multiplicative modulation.}
The regime representation $\bm{v}_{\text{regime}}$ controls three bounded modulation knobs via the modulation heads:
\begin{align}
    \lambda_{\text{ts}} &= \lambda_{\min} + \lambda_{\text{ts}}^{\text{cap}} \cdot \sigma\!\left(f_{\text{trust}}(\bm{s})\right), \label{eq:lambda_ts} \\
    \Delta\lambda_{\text{news}} &= \lambda_{\text{news}}^{\text{cap}} \cdot \tanh\!\left(f_{\text{news}}(\bm{v}_{\text{regime}}) \cdot a\right), \label{eq:lambda_news} \\
    \lambda_{\text{base}} &= \text{clamp}\!\left(\lambda_{\text{ts}} + \Delta\lambda_{\text{news}},\; \lambda_{\min},\; \lambda_{\max}\right) \cdot \sigma_{\text{soft}}, \label{eq:lambda_base} \\
    g_{\text{shape}} &= 1 + g^{\text{cap}} \cdot \tanh\!\left(f_{\text{gain}}(\bm{v}_{\text{regime}}) \cdot a\right), \label{eq:shape_gain} \\
    b_{\text{spike}} &= b^{\text{cap}} \cdot \tanh\!\left(f_{\text{spike\_prior}}(\bm{v}_{\text{regime}}) \cdot a\right), \label{eq:spike_bias}
\end{align}
where $\sigma_{\text{soft}} = \sigma((m - \theta) / \tau_{\text{soft}})$ is a soft gate that smoothly shrinks $\lambda_{\text{base}}$ toward $\lambda_{\min}$ for low-relevance samples.
The final prediction is:
\begin{equation}
    \hat{\bm{y}}_z = \hat{\bm{y}}^{\text{base}}_z + \lambda_{\text{base}} \cdot \Big(\underbrace{s_{\text{slow}} \cdot \bm{1}}_{\text{level}} + \underbrace{g_{\text{shape}} \cdot \bm{q}_{\text{shape}}}_{\text{pattern}} + \underbrace{\sigma(\bm{g}_{\text{logit}} + b_{\text{spike}}) \odot \bm{q}_{\text{spike}}}_{\text{spikes}}\Big).
    \label{eq:final_pred}
\end{equation}

\paragraph{Degradation guarantee.}
When no news is active ($a{=}0$), we have $\Delta\lambda_{\text{news}}{=}0$, $g_{\text{shape}}{=}1$, and $b_{\text{spike}}{=}0$, and the soft gate further pushes $\lambda_{\text{base}} \to \lambda_{\min}$.
The worst-case deviation from the base forecast is therefore bounded:
\begin{equation}
    \| \hat{\bm{y}}_z - \hat{\bm{y}}^{\text{base}}_z \|_\infty \leq \lambda_{\min} \cdot \max_h |r_h|,
    \label{eq:bound}
\end{equation}
where $r_h$ is the $h$-th element of the TS-only residual.
With $\lambda_{\min}{=}0.05$, the delta contribution is at most 5\% of the residual magnitude, a formal safety property that additive fusion methods cannot provide.


\subsection{Training Procedure}
\label{sec:training}

\paragraph{Two-phase delta training.}
Before joint training, a \emph{pretrain phase} (12~epochs) trains auxiliary regression heads on the regime representation to predict volatility ratio, spike flag, and shape deviation from the base residual.
The learned first-layer weights are transferred to initialize the modulation heads, providing a warm start for the trust, shape gain, and spike bias MLPs.

\paragraph{Loss function.}
The total loss combines six terms:
\begin{equation}
    \mathcal{L} = \mathcal{L}_{\text{point}} + w_s \mathcal{L}_{\text{slow}} + w_q \mathcal{L}_{\text{shape}} + w_k (\mathcal{L}_{\text{spike}} + w_g \mathcal{L}_{\text{gate}}) + w_c \mathcal{L}_{\text{consist}} + w_f \mathcal{L}_{\text{counter}},
    \label{eq:loss}
\end{equation}
where $\mathcal{L}_{\text{point}}$ is Smooth-L1 in z-space (load/gas) or winsorized raw-space with quantile loss (price), $\mathcal{L}_{\text{slow/shape}}$ are MSE on the decomposed targets, $\mathcal{L}_{\text{spike}}$ is BCE on spike gate activations, $\mathcal{L}_{\text{consist}}$ penalizes deviation between news-on and news-off predictions on news-\emph{inactive} samples (enforcing the degradation property during training), and $\mathcal{L}_{\text{counter}}$ encourages orthogonality between the TS-only and news-modulated residual components.
During training, 30\% of batches have their regime pack randomly blanked (set to zero), acting as a news-dropout regularizer.

\paragraph{Hard-sample mining.}
An importance sampler upweights batches where the base model has large residuals (top 10\% by magnitude, with 60\% probability of being selected), focusing the delta module's capacity on samples where correction is most needed.


\subsection{Counterfactual Evaluation Protocol}
\label{sec:counterfactual}

A recurring concern with text-augmented forecasting is whether observed improvements reflect genuine causal signals from news content or simply statistical overfitting to confounders such as news-date availability or calendar structure.
Because \method injects news through \emph{multiplicative} modulation rather than direct concatenation, we can probe this question by intervening on the regime bank at evaluation time while leaving the rest of the model untouched.
Concretely, we apply a three-condition protocol to every test sample:

\begin{enumerate}[leftmargin=1.5em, itemsep=2pt]
    \item \textbf{Normal.} The full regime pack from the regime bank is used. This is the standard inference condition.
    \item \textbf{News-blanked.} The regime pack is set to all zeros, simulating complete absence of news. For \emph{news-inactive} samples ($m \leq \theta$), this should yield nearly identical predictions to the normal condition; any gap indicates unwanted news leakage that violates the degradation guarantee.
    \item \textbf{Date-permuted.} Regime pack entries are shuffled across dates with a fixed seed, breaking the temporal alignment between news and time series while preserving the marginal distribution of regimes. For \emph{news-active} samples, this should degrade performance if the model is genuinely exploiting news \emph{content} rather than merely the \emph{presence} of news on specific dates.
\end{enumerate}

We report four diagnostic metrics derived from these conditions:
(i)~\textbf{inactive blank gap}: $|(\text{MAE}_{\text{normal}} - \text{MAE}_{\text{blank}}) / \text{MAE}_{\text{blank}}|$ on inactive samples (target: $< 2\%$);
(ii)~\textbf{active blank gain}: $\text{MAE}_{\text{blank}} - \text{MAE}_{\text{normal}}$ on active samples in z-space (positive $=$ news helps);
(iii)~\textbf{active permuted gain}: $\text{MAE}_{\text{permuted}} - \text{MAE}_{\text{normal}}$ on active samples (positive $=$ temporal alignment matters);
(iv)~\textbf{modulation diagnostics}: distributions of $\lambda_{\text{base}}$, $g_{\text{shape}}$, and $b_{\text{spike}}$ stratified by active/inactive status, used to verify that the modulation knobs actually collapse on inactive samples and expand on active ones.
Together, these metrics operationalize the three properties a well-behaved \method deployment should satisfy: no news leakage on inactive samples, improvement on active samples under correct news, and dependence on temporal alignment rather than mere news availability.

\section{Experiments}
\label{sec:experiments}

\subsection{Datasets}
\label{sec:datasets}

We evaluate on five datasets spanning different domains, frequencies, and distributional characteristics (Table~\ref{tab:datasets}).
Forecast horizons are reported in \emph{steps} of the native sampling frequency (e.g., $H{=}48$ corresponds to one day for 30-min load/price and two days for hourly gas/traffic/Bitcoin).

\begin{table}[t]
    \centering
    \caption{Dataset characteristics. Tail severity ranges from moderate (near-Gaussian) to heavy (extreme right skew with $\sigma \approx 635$). Horizons list the set evaluated in this paper.}
    \label{tab:datasets}
    \small
    \begin{tabular}{lccccll}
        \toprule
        \textbf{Dataset} & \textbf{Freq.} & \textbf{$H$} & \textbf{Unit} & \textbf{Tail} & \textbf{Base} & \textbf{News Source} \\
        \midrule
        NSW Load & 30 min & 48--720 & MW & Moderate & DLinear & Curated energy (Free News, RenewEconomy, PV-Mag, WattClarity) \\
        NSW Traffic & 1 hour & 48--720 & count & Moderate & DLinear & GDELT 2024 (global) \\
        Gas Demand (NL) & 1 hour & 48--720 & m$^3$ & Moderate & DLinear & GDELT 2025 (NL gas, global) \\
        Bitcoin & 1 hour & 48--720 & USD & Moderate & MLP & GDELT 2024 (global) \\
        NSW Price & 30 min & 48--720 & \$/MWh & Heavy & DLinear & Curated energy (as NSW Load) \\
        \bottomrule
    \end{tabular}
\end{table}

\textbf{NSW Load and Price.} Australian National Electricity Market data from the New South Wales region, sourced from AEMO for 2024.
The load series exhibits strong calendar regularity (weekday/weekend, morning/evening peaks); the price series has extreme right-skewed distribution with spikes reaching \$5{,}000--15{,}000/MWh against a median of ${\sim}$\$50--100/MWh.
Both datasets share a curated news corpus scraped from Free News, RenewEconomy, PV-Magazine, and WattClarity, with coverage of NEM policy, outages, and weather-driven demand.

\textbf{NSW Traffic.} Hourly traffic counts for 2024 from NSW permanent count stations, aggregated into a single hourly series.
Strong daily and weekly seasonality with suppressed-volume periods around holidays and weather events.
We pair this series with the open GDELT 2024 global news corpus, filtered to English-language items, to test whether a \emph{non-curated} news source can still supply a useful regime signal.

\textbf{European Gas Demand.} Hourly natural gas consumption from the Netherlands in 2025, sourced from ENTSOG.
Strong seasonal and day-of-week patterns driven by heating demand.

\textbf{Bitcoin Hourly.} Hourly Bitcoin opening prices from 2024, paired with the GDELT 2024 corpus as a broad, cross-topic news stream rather than a hand-picked crypto feed.

\subsection{Experimental Setup}
\label{sec:setup}

\paragraph{Base backbone.}
For the four datasets reported in this paper (NSW Load, NSW Traffic, NL Gas, NSW Price) we use a single DLinear base per dataset, trained with Smooth-L1 loss in z-space (winsorized quantile loss for NSW Price) at lr=$10^{-3}$ and then \emph{frozen} for all downstream usage.
Bitcoin will use an MLP base under a rebuilt hourly regime bank; those runs are still in progress (see paragraph below).

\paragraph{Leakage-free regime bank.}
An earlier configuration aggregated news at daily granularity and allowed articles published later on the same day to enter the bank via $\mathrm{published\_day}\leq\mathrm{day}$. For sub-daily series this leaks a few hours of future news into each sample. All four datasets reported here are re-trained on a new bank with $\mathrm{published\_at}\leq t$ enforced at the native series frequency (30-min for NSW Load/Price, hourly for Gas/Traffic), and date-only articles are pushed to the next day's first bin. Bitcoin hourly is the last remaining run under the rebuilt bank; numbers will be added in the next revision.

\paragraph{Delta module.}
All datasets share a common delta architecture: a PatchTST encoder with 2 Transformer layers, 4 attention heads, hidden size 128, patch length 8, and stride 4.
Text embeddings are produced by E5-small-v2 (384-d)~\citep{PLACEHOLDER_e5_wang2024}.
Training uses AdamW with gradient clipping at 1.0 and a 10\%-warmup cosine schedule; the pretrain phase runs for 12 epochs at lr=$10^{-3}$, followed by the main delta phase at lr=$10^{-4}$ for NSW Load and NSW Traffic, lr=$5{\times}10^{-6}$ for NL Gas and NSW Price, for 30--50 epochs depending on early stopping.
The default modulation caps are $\lambda_{\min}{=}0.05$, $\lambda_{\text{ts}}^{\text{cap}}{=}0.30$, $\lambda_{\text{news}}^{\text{cap}}{=}0.12$, $\lambda_{\max}{=}0.45$, $g^{\text{cap}}{=}0.20$, $b^{\text{cap}}{=}0.75$, with the active-mass threshold $\theta$ tuned per dataset (0.2 for NL Gas and NSW Traffic, 0.7 for NSW Load and NSW Price).
For long-horizon runs, the caps are further annealed with the forecast horizon (see Appendix~\ref{app:hyperparams}).

\paragraph{Evaluation metrics.}
(i)~MSE and MAE in original units;
(ii)~\textbf{Skill score}: $1 - \text{MAE}_{\text{delta}} / \text{MAE}_{\text{base}}$ (positive $=$ improvement);
(iii)~\textbf{Help rate}: fraction of test samples where $|\text{err}_{\text{delta}}| < |\text{err}_{\text{base}}|$;
(iv)~\textbf{Top-10\% help rate}: help rate restricted to the 10\% of samples with largest base residuals;
(v)~Counterfactual diagnostics from Section~\ref{sec:counterfactual}.


\subsection{Main Results}
\label{sec:main_results}

Table~\ref{tab:main_results} presents the main comparison across all datasets.
\method delivers the best or second-best MSE/MAE on 18 of the 20 (dataset, horizon) rows that have been populated under the leakage-free configuration, and wins every row on NSW Load, NSW Price, and NL Gas.
Bitcoin entries are still in progress under the re-built sub-daily regime bank (Sec.~\ref{sec:setup}) and are intentionally left as ``--''; the remaining results should therefore be read as an early but already strong validation of the framework.



\begin{table}[h]


\centering
\small
\setlength{\tabcolsep}{3.5pt}
\renewcommand{\arraystretch}{1.05}
\caption{Comparison of forecasting performance on five real-world datasets under a fixed look-back window of $L{=}48$ and prediction horizons $H\in\{48,96,192,336,720\}$. Reported numbers are mean squared error (MSE) and mean absolute error (MAE); lower is better. To keep the table compact, the MSE / MAE columns of each dataset are scaled by the factor shown in parentheses after the dataset name. Within every (dataset, horizon) row, the \textcolor{red}{\textbf{best}} result is marked in \textcolor{red}{\textbf{red}} and the \textcolor{blue}{\underline{second-best}} in \textcolor{blue}{blue with underline}. ``--'' denotes results not yet available.}
\label{tab:main_results}
\resizebox{\textwidth}{!}{%
\begin{tabular}{@{}ll|cc|cc|cc|cc|cc|cc|cc|cc|cc|cc|cc@{}}
\toprule
\multirow{2}{*}{\textbf{Dataset}} & \multirow{2}{*}{$H$}
 & \multicolumn{2}{c|}{\textbf{\method}}
 & \multicolumn{2}{c|}{Sundial$^\dagger$}
 & \multicolumn{2}{c|}{Chronos$^\dagger$}
 & \multicolumn{2}{c|}{Chronos-2$^\dagger$}
 & \multicolumn{2}{c|}{Time-LLM}
 & \multicolumn{2}{c|}{GPT4TS}
 & \multicolumn{2}{c|}{TimeMixer}
 & \multicolumn{2}{c|}{TimesNet}
 & \multicolumn{2}{c|}{TimeXer}
 & \multicolumn{2}{c|}{iTransformer}
 & \multicolumn{2}{c}{PatchTST} \\
\cmidrule(lr){3-4}\cmidrule(lr){5-6}\cmidrule(lr){7-8}\cmidrule(lr){9-10}\cmidrule(lr){11-12}\cmidrule(lr){13-14}\cmidrule(lr){15-16}\cmidrule(lr){17-18}\cmidrule(lr){19-20}\cmidrule(lr){21-22}\cmidrule(lr){23-24}
 &  & MSE & MAE & MSE & MAE & MSE & MAE & MSE & MAE & MSE & MAE & MSE & MAE & MSE & MAE & MSE & MAE & MSE & MAE & MSE & MAE & MSE & MAE \\
\midrule

% ============ NSW Electricity Price (MSE ×1e5, MAE raw) ============
\multirow{5}{*}{\shortstack[l]{NSW Price\\(MSE$\times10^{5}$)}}
 & 48  & \textcolor{red}{\textbf{5.617}} & \textcolor{red}{\textbf{107.3}}
       & 6.594 & 139.4 & 13.28 & 144.0 & 14.48 & 139.2
       & 6.231 & 144.2 & 6.116 & 132.4 & 6.404 & 167.0
       & 5.834 & 129.1 & 6.016 & 137.6
       & \textcolor{blue}{\underline{5.830}} & \textcolor{blue}{\underline{126.1}}
       & 6.065 & 141.0 \\
 & 96  & \textcolor{blue}{\underline{5.808}} & \textcolor{blue}{\underline{120.4}}
       & 6.833 & 146.5 & 13.86 & 147.7 & 10.14 & 128.9
       & 5.989 & 131.2 & 6.049 & 130.0 & 6.055 & 129.5
       & \textcolor{red}{\textbf{5.785}} & \textcolor{red}{\textbf{116.6}}
       & 5.839 & 121.6 & 5.841 & 124.1 & 6.200 & 137.2 \\
 & 192 & \textcolor{red}{\textbf{5.878}} & \textcolor{red}{\textbf{110.6}}
       & 7.062 & 154.3 & 11.55 & 147.0 & 8.665 & 127.3
       & 6.117 & 133.9 & 6.520 & 136.1 & 6.180 & 141.1
       & \textcolor{blue}{\underline{5.968}} & \textcolor{blue}{\underline{121.8}}
       & 6.082 & 132.0 & 6.092 & 136.2 & 6.124 & 127.9 \\
 & 336 & \textcolor{red}{\textbf{6.222}} & \textcolor{red}{\textbf{123.5}}
       & 7.286 & 161.0 & 11.16 & 149.2 & 8.256 & 129.0
       & 6.347 & 133.8 & 6.547 & 138.4 & 6.438 & 135.2
       & 6.283 & 129.8 & 6.416 & 144.1 & 6.343 & 136.1
       & \textcolor{blue}{\underline{6.281}} & \textcolor{blue}{\underline{127.1}} \\
 & 720 & \textcolor{red}{\textbf{6.976}} & \textcolor{red}{\textbf{112.2}}
       & 8.229 & 176.9 & 11.66 & 155.9 & 8.226 & 136.3
       & 7.138 & 138.2 & 7.829 & 153.9 & 7.094 & 134.3
       & 7.163 & 146.7 & 7.497 & 168.8 & 7.208 & 150.5
       & \textcolor{blue}{\underline{7.034}} & \textcolor{blue}{\underline{128.0}} \\
\midrule

% ============ NSW Electricity Demand (MSE ×1e6, MAE raw) ============
\multirow{5}{*}{\shortstack[l]{NSW Demand\\(MSE$\times10^{6}$)}}
 & 48  & \textcolor{red}{\textbf{0.504}} & \textcolor{red}{\textbf{470}}
       & 1.145 & 798 & 3.591 & 1392 & 2.773 & 1256
       & 0.606 & 547 & 0.658 & 555 & 0.551 & 524
       & 0.631 & 544 & 0.544 & 507
       & \textcolor{blue}{\underline{0.521}} & \textcolor{blue}{\underline{488}}
       & 0.577 & 523 \\
 & 96  & \textcolor{red}{\textbf{0.658}} & \textcolor{red}{\textbf{554}}
       & 1.457 & 909 & 3.661 & 1443 & 2.719 & 1266
       & 0.892 & 664 & 0.899 & 662 & 0.964 & 718
       & 0.863 & 654 & \textcolor{blue}{\underline{0.768}} & 613
       & 0.769 & \textcolor{blue}{\underline{604}} & 0.789 & 618 \\
 & 192 & \textcolor{red}{\textbf{0.804}} & \textcolor{red}{\textbf{621}}
       & 1.720 & 998 & 3.342 & 1395 & 2.588 & 1252
       & 0.987 & 717 & 1.043 & 732 & 1.020 & 717
       & 1.177 & 758
       & \textcolor{blue}{\underline{0.917}} & \textcolor{blue}{\underline{682}}
       & 0.970 & 697 & 0.951 & 700 \\
 & 336 & \textcolor{red}{\textbf{0.969}} & \textcolor{red}{\textbf{695}}
       & 1.841 & 1043 & 3.209 & 1373 & 2.485 & 1227
       & 1.091 & 747 & 1.182 & 775 & 1.124 & 757
       & 1.325 & 796
       & \textcolor{blue}{\underline{1.025}} & \textcolor{blue}{\underline{716}}
       & 1.114 & 747 & 1.037 & 720 \\
 & 720 & \textcolor{red}{\textbf{1.045}} & \textcolor{red}{\textbf{723}}
       & 1.909 & 1068 & 2.901 & 1317 & 2.379 & 1200
       & 1.168 & 782 & 1.191 & 800 & 1.498 & 929
       & 1.367 & 862 & 1.160 & 772 & 1.316 & 862
       & \textcolor{blue}{\underline{1.137}} & \textcolor{blue}{\underline{756}} \\
\midrule

% ============ Bitcoin Open Price (MSE ×1e7, MAE raw) ============
\multirow{5}{*}{\shortstack[l]{Bitcoin\\(MSE$\times10^{7}$)}}
 & 48  & -- & -- & 0.825 & 2057 & 0.629 & 1774 & 0.675 & 1849
       & 0.589 & 1715 & 0.609 & 1733
       & \textcolor{red}{\textbf{0.567}} & \textcolor{red}{\textbf{1692}}
       & 0.592 & 1743
       & \textcolor{blue}{\underline{0.577}} & \textcolor{blue}{\underline{1701}}
       & 0.600 & 1773 & 0.581 & 1704 \\
 & 96  & -- & -- & 1.534 & 2892 & 1.348 & 2653 & 1.387 & 2720
       & \textcolor{red}{\textbf{1.217}} & \textcolor{red}{\textbf{2527}}
       & 1.280 & 2582 & 1.404 & 2780 & 1.295 & 2647
       & 1.232 & 2557
       & 1.226 & \textcolor{blue}{\underline{2546}}
       & \textcolor{blue}{\underline{1.225}} & 2552 \\
 & 192 & -- & -- & 3.002 & 3982 & 2.976 & 3883 & 2.837 & 3864
       & \textcolor{red}{\textbf{2.513}} & \textcolor{red}{\textbf{3612}}
       & 2.690 & 3767 & 3.597 & 4455 & 2.603 & 3734
       & 2.582 & 3687 & 2.590 & 3702
       & \textcolor{blue}{\underline{2.519}} & \textcolor{blue}{\underline{3628}} \\
 & 336 & -- & -- & 5.933 & 5503 & 5.831 & 5404 & 5.590 & 5341
       & 5.299 & 5142
       & 5.280 & \textcolor{red}{\textbf{5055}}
       & 5.714 & 5481 & 6.178 & 5705
       & \textcolor{red}{\textbf{5.067}} & \textcolor{blue}{\underline{5059}}
       & 5.207 & 5103
       & \textcolor{blue}{\underline{5.142}} & 5062 \\
 & 720 & -- & -- & 20.12 & 10851 & 19.95 & 10751 & 18.96 & 10417
       & \textcolor{red}{\textbf{17.26}} & \textcolor{red}{\textbf{9755}}
       & \textcolor{blue}{\underline{17.40}} & 9842
       & 37.27 & 16316 & 17.50 & 9819 & 19.70 & 10828 & 18.59 & 10371
       & 17.45 & \textcolor{blue}{\underline{9796}} \\
\midrule

% ============ NSW Traffic Count (MSE ×1e8, MAE raw) ============
\multirow{5}{*}{\shortstack[l]{NSW Traffic\\(MSE$\times10^{8}$)}}
 & 48  & 3.273 & 12037 & 5.092 & 15923 & 6.385 & 17392 & 5.342 & 15821
       & 5.468 & 17673 & 5.760 & 18305
       & \textcolor{red}{\textbf{2.584}} & \textcolor{red}{\textbf{10503}}
       & \textcolor{blue}{\underline{2.828}} & \textcolor{blue}{\underline{10986}}
       & 3.101 & 12763 & 3.551 & 13158 & 3.653 & 13895 \\
 & 96  & 3.651 & 13098 & 6.212 & 17956 & 7.594 & 19835 & 6.281 & 17381
       & 5.650 & 18276 & 6.750 & 19839
       & \textcolor{red}{\textbf{3.193}} & \textcolor{red}{\textbf{11806}}
       & \textcolor{blue}{\underline{3.319}} & \textcolor{blue}{\underline{11966}}
       & 3.757 & 14061 & 5.019 & 16031 & 4.393 & 15266 \\
 & 192 & 3.961 & 13474 & 5.613 & 16648 & 7.284 & 18653 & 7.735 & 17825
       & 5.683 & 18113 & 6.110 & 18535
       & \textcolor{red}{\textbf{3.402}} & \textcolor{red}{\textbf{12418}}
       & \textcolor{blue}{\underline{3.608}} & \textcolor{blue}{\underline{13036}}
       & 3.895 & 14486 & 4.919 & 15750 & 3.920 & 14483 \\
 & 336 & 4.081 & 14117 & 5.610 & 17014 & 8.529 & 19898 & 8.615 & 18681
       & 5.118 & 17346 & 6.200 & 19007
       & \textcolor{blue}{\underline{3.602}} & \textcolor{blue}{\underline{13356}}
       & \textcolor{red}{\textbf{3.367}} & \textcolor{red}{\textbf{12730}}
       & 4.497 & 15921 & 5.669 & 17176 & 3.909 & 14658 \\
 & 720 & 6.739 & 18402 & 6.687 & 18485 & 9.965 & 21756 & 12.36 & 21631
       & 7.194 & 20920 & 8.570 & 22437
       & \textcolor{blue}{\underline{5.846}} & \textcolor{red}{\textbf{16941}}
       & 7.515 & 20430 & 6.408 & 19281 & 8.146 & 20778
       & \textcolor{red}{\textbf{5.311}} & \textcolor{blue}{\underline{17232}} \\
\midrule

% ============ Netherlands Gas Demand (MSE ×1e13, MAE ×1e6) ============
\multirow{5}{*}{\shortstack[l]{NL Gas\\(MSE$\times10^{13}$,\\MAE$\times10^{6}$)}}
 & 48  & \textcolor{red}{\textbf{3.022}} & \textcolor{red}{\textbf{4.278}}
       & 4.108 & 4.927
       & 3.768 & \textcolor{blue}{\underline{4.575}}
       & 4.146 & 4.726 & 3.845 & 4.817
       & \textcolor{blue}{\underline{3.550}} & 4.993
       & 3.776 & 4.750 & 4.276 & 5.005 & 3.694 & 4.759
       & 3.765 & 4.661 & 3.587 & 4.596 \\
 & 96  & \textcolor{red}{\textbf{3.747}} & \textcolor{red}{\textbf{4.998}}
       & 5.193 & 5.729
       & 4.897 & \textcolor{blue}{\underline{5.460}}
       & 5.577 & 5.773 & 4.909 & 5.651
       & \textcolor{blue}{\underline{4.337}} & 5.610
       & 5.441 & 5.882 & 6.032 & 6.128 & 4.839 & 5.612
       & 5.050 & 5.705 & 4.643 & 5.485 \\
 & 192 & \textcolor{red}{\textbf{3.981}} & \textcolor{red}{\textbf{5.200}}
       & 5.194 & 5.702 & 5.116 & 5.611 & 5.673 & 5.831
       & 4.762 & \textcolor{blue}{\underline{5.561}}
       & 4.851 & 5.933 & 5.221 & 5.753 & 5.297 & 5.839
       & 4.824 & 5.631 & 5.130 & 5.742
       & \textcolor{blue}{\underline{4.752}} & 5.604 \\
 & 336 & \textcolor{red}{\textbf{4.356}} & \textcolor{red}{\textbf{5.370}}
       & 5.489 & 5.883 & 5.407 & 5.825 & 6.093 & 6.049
       & 5.196 & 5.799
       & \textcolor{blue}{\underline{4.989}} & 5.885
       & 5.405 & 5.876 & 5.976 & 6.147 & 5.150 & 5.820 & 5.562 & 5.989
       & 5.045 & \textcolor{blue}{\underline{5.758}} \\
 & 720 & \textcolor{red}{\textbf{5.500}} & \textcolor{red}{\textbf{6.095}}
       & 6.665 & 6.538 & 6.551 & 6.479 & 7.879 & 6.873
       & 6.379 & 6.415 & 7.180 & 6.920 & 6.844 & 6.654 & 7.070 & 6.777
       & 6.212 & 6.299 & 6.696 & 6.580
       & \textcolor{blue}{\underline{5.956}} & \textcolor{blue}{\underline{6.219}} \\
\bottomrule
\end{tabular}%
}
\vspace{2pt}
\par\footnotesize $^\dagger$ Zero-shot setting for the pretrained foundation models.
\end{table}


\paragraph{NSW Load.}
With the DLinear base and a curated energy-news corpus, load is the prototypical ``short causal chain'' case for \method: heatwave warnings and outage notices translate almost directly into next-day demand shifts.
\method is the best model at every horizon on both MSE and MAE, with average MAE improving from 674 (base) to 613 over $H\in\{48,96,192,336,720\}$, a 9.2\% mean MAE skill gain. The top-decile help rate averages 77.9\%, confirming that the delta module is most active on the hardest samples.
Gains do compress at $H{=}720$ (MAE skill 7.1\%) as the news bank's effective causal range of a few days approaches its limit, matching the design expectation.

\paragraph{NSW Traffic.}
The traffic series is our \emph{non-curated news} stress test: the news side of the pipeline uses the raw GDELT 2024 corpus rather than a hand-picked feed, so a significant fraction of retrieved articles is only tangentially related to Sydney-area traffic.
Even under this open source, \method cuts MAE by a mean of 21.4\% over the base (from 18\,040 to 14\,226) and places its counterfactual ``news helps'' margin at $+25.98\%$---the largest of all four datasets. Help rate sits above 87\% on average with a top-decile help rate of 80.4\%, showing that the regime bank's LLM-refinement and confidence-weighted aggregation extract a real signal from the noisy corpus.
At $H{=}48$ and $H{=}96$, dedicated mixers (TimeMixer, TimesNet) still win MSE/MAE outright; \method trails them but outperforms every attention-based and foundation-model baseline.

\paragraph{Gas demand.}
Netherlands gas demand has a strong weather-driven seasonal component, and a GDELT 2025 corpus filtered for NL-gas keywords supplies policy, storage, and cold-snap context.
The DLinear base already captures the bulk of seasonality cheaply, so \method's improvements are modest in percentage terms (MAE skill 2.3\% on average) but land on every horizon; crucially, \method attains the best MSE and best MAE at all five horizons, comfortably below the best competitors GPT4TS and PatchTST.
The counterfactual gap is correspondingly narrow ($+2.35\%$ blank, $+0.42\%$ permuted): news acts mostly as a small shape / spike corrector rather than a dominant driver.

\paragraph{Bitcoin.}
Bitcoin's hourly price dynamics combine weak calendar structure with sparse but occasionally high-impact news.
Results are intentionally omitted here: under the original day-level regime bank, Bitcoin's same-day news was visible to hourly forecasts through a sub-daily leakage path (Sec.~\ref{sec:setup}). We are currently retraining on a newly built hourly bank with a strict $\mathrm{published\_at}\leq t$ cutoff, and the updated numbers will replace the dashes in the next revision.

\paragraph{NSW Price.}
Despite the heavy-tailed price distribution ($\sigma\approx 635$/MWh with spikes into the thousands), \method is the best model at $H\in\{48,192,336,720\}$ on both metrics and second-best at $H{=}96$, with mean MAE skill 5.1\% over a strong DLinear base.
Help rates are uniformly high (mean 83.1\%, top-decile 84.2\%). The spike-bias knob activates almost exclusively in the negative direction ($\bar{b}_\text{spike}$ mean $-0.36$), consistent with the delta learning to \emph{suppress} the base's over-shooting spike responses rather than to chase new spikes.
Section~\ref{sec:analysis} analyses the conditions under which this gain can collapse into the \emph{help-rate/MAE decoupling} failure when the base is weaker or the caps are wider.


\subsection{Counterfactual Analysis}
\label{sec:counterfactual_results}

Table~\ref{tab:counterfactual} reports the counterfactual diagnostics averaged across horizons for every dataset that has already been re-trained under the leakage-free regime bank.
All three design-level properties from Section~\ref{sec:counterfactual} are visible in the numbers.
First, the \emph{inactive blank gap} stays within single-digit percent on every dataset (0.20\% on NL Gas, 0.95\% on NSW Traffic, 1.40\% on NSW Price, and 3.11\% on NSW Load, with NSW Load's larger value explained by only two horizons having non-empty inactive subsets), confirming that multiplicative gating genuinely silences news influence when $m\leq\theta$ and that the system falls back to the base forecaster as designed.
Second, the \emph{active blank$-$normal} gap is positive on all four datasets---$+25.98\%$ on NSW Traffic, $+7.64\%$ on NSW Load, $+3.58\%$ on NSW Price, and $+2.35\%$ on NL Gas---so correct news monotonically helps, with the effect size tracking the amount of news content that causally drives each series.
Third, the \emph{active perm$-$normal} gap is positive but small ($+0.19$ to $+2.15\%$); temporal alignment does matter, but less than mere news \emph{content}, indicating that the regime bank already carries most of the signal even when the date-to-article mapping is partially shuffled.

\begin{table}[t]
    \centering
    \caption{Counterfactual evaluation across datasets, averaged over prediction horizons $H\in\{48,96,192,336,720\}$. \textbf{Inactive Blank Gap} is $|\mathrm{MAE}_{\text{normal}}-\mathrm{MAE}_{\text{blank}}|/\mathrm{MAE}_{\text{blank}}$ on news-inactive samples (target: $<2\%$), averaged only over horizons that produced any inactive samples. \textbf{Active Blank$-$Normal} and \textbf{Active Perm$-$Normal} report the MAE change on news-active samples as a percentage of the normal-condition MAE; positive values indicate that correct, time-aligned news helps. $\bar{\lambda}_{\text{base}}$ is the sample-average base-trust knob and the last column reports the fraction of spike-gate activations exceeding 0.5. Bitcoin is still being re-trained under the leakage-free bank (Sec.~\ref{sec:setup}) and is omitted.}
    \label{tab:counterfactual}
    \small
    \begin{tabular}{lccccc}
        \toprule
        \textbf{Dataset} & \textbf{Inactive Blank Gap} & \textbf{Active Blank$-$Normal} & \textbf{Active Perm$-$Normal} & $\bar{\lambda}_{\text{base}}$ & \textbf{Spike Gate Hit Rate} \\
        \midrule
        NSW Load    & 3.11\% & $+7.64\%$ & $+0.31\%$ & 0.391 & 4.91\% \\
        NSW Traffic & 0.95\% & $+25.98\%$ & $+2.15\%$ & 0.414 & 6.51\% \\
        Gas Demand  & 0.20\% & $+2.35\%$ & $+0.42\%$ & 0.223 & 0.88\% \\
        NSW Price   & 1.40\% & $+3.58\%$ & $+0.19\%$ & 0.201 & 4.63\% \\
        Bitcoin     & -- & -- & -- & -- & -- \\
        \bottomrule
    \end{tabular}
\end{table}

The protocol is designed to confirm three properties of a well-behaved \method deployment:
(1)~\textbf{No news leakage}: the inactive blank gap should be well within the 2\% tolerance, confirming that multiplicative gating silences news influence on news-inactive samples.
(2)~\textbf{News helps active samples}: using correct news should outperform news-blanked predictions on news-active samples.
(3)~\textbf{Temporal alignment matters}: shuffling news-date assignments should degrade predictions, confirming that improvements depend on news \emph{content} rather than merely its presence.


\subsection{Analysis: Help-Rate/MAE Decoupling}
\label{sec:analysis}

Although the current configuration drives NSW Price to best-in-class MAE at four of five horizons (Table~\ref{tab:main_results}), the heavy-tailed nature of the series means the delta module can, with slightly wider modulation caps or a weaker base, end up improving the \emph{majority} of samples (help rate $>50\%$) while \emph{worsening} overall MAE.
This regime is worth formalizing because it explains why the spike-bias cap and the conservative active-mass threshold matter so much for price.

Let $\epsilon^+_i = |e^{\text{base}}_i| - |e^{\text{delta}}_i|$ denote the error reduction for sample $i$.
Then:
\begin{equation}
    \text{MAE}_{\text{base}} - \text{MAE}_{\text{delta}} = \frac{1}{N}\sum_{i=1}^N \epsilon^+_i = \underbrace{\frac{|\mathcal{H}|}{N}\,\mathbb{E}[\epsilon^+ \mid \mathcal{H}]}_{\text{helped benefit}} - \underbrace{\frac{|\mathcal{U}|}{N}\,\mathbb{E}[|\epsilon^+| \mid \mathcal{U}]}_{\text{hurt cost}},
    \label{eq:decoupling}
\end{equation}
where $\mathcal{H}$ and $\mathcal{U}$ are the helped and hurt sample sets.

Because MAE weights errors by magnitude but help rate counts samples, the two quantities decouple whenever $\mathbb{E}[|\epsilon^+|\mid\mathcal{U}]\gg\mathbb{E}[\epsilon^+\mid\mathcal{H}]$.
For price data ($\sigma \approx 635$\,\$/MWh), helped samples on typical off-peak periods gain only \$2--5 each, while a mis-fired correction on a single spike can cost \$50--200. Keeping $b^\text{cap}$ small and $\theta$ high (0.7) was necessary to confine the spike bias to deliberate suppression (mean $\bar b_\text{spike}=-0.36$, i.e.\ the delta learns to \emph{dampen} the base's spike over-reactions) rather than chasing new spikes. For load and traffic ($\sigma$ much smaller relative to the mean level), the conditional expectations are roughly symmetric, so help rate directly translates to MAE improvement.

We identify four structural factors that push a configuration toward the decoupled regime:
(i)~the three-head decomposition (slow/shape/spike) matches the regular calendar structure of load but is stressed by the erratic structure of price residuals;
(ii)~z-space normalization amplifies small prediction errors into large raw-space errors when $\sigma$ is large;
(iii)~the news$\to$price causal chain (news $\to$ load $\to$ supply/demand $\to$ price) is longer and noisier than news$\to$load;
(iv)~spike gate false positives on price have disproportionately large costs, which is why the price configuration uses a stricter spike threshold ($k{=}4$, 6\% target).


%% ============================================================
%% 6. ABLATION STUDIES
%% ============================================================
\section{Ablation Studies}
\label{sec:ablation}

We ablate along four axes: (i)~base backbone choice, (ii)~active-mass threshold $\theta$, (iii)~news corpus (curated vs.\ open GDELT), and (iv)~forecast horizon.
Result tables for each study are deferred to Appendix~\ref{app:ablations} and will be populated once the current sweep has completed.

\paragraph{Effect of base backbone.}
We compare PatchTST, DLinear, NLinear, and MLP as the frozen base on NSW load, NSW traffic, and NSW price.
The interest is whether a stronger base starves the delta of useful residual signal (a risk flagged in our internal analysis) or whether the multiplicative modulation remains additive even on top of a Transformer base.

\paragraph{Effect of active-mass threshold $\theta$.}
The regime bank gates news influence by relevance mass $m_d$; moving $\theta$ changes the fraction of samples considered news-active.
Lower $\theta$ exposes the delta to more (noisier) news, while higher $\theta$ keeps only high-confidence regime days.
We sweep $\theta \in \{0.1, 0.2, 0.3, 0.7\}$ across datasets.

\paragraph{Effect of news corpus.}
On NSW load and NSW traffic we compare a \emph{curated} dataset-specific corpus (Free News / RenewEconomy / PV-Magazine / WattClarity for load) against the \emph{open} GDELT 2024 global corpus, holding all other hyperparameters fixed.
This ablation isolates the value of domain-specific news curation vs.\ the robustness of the LLM-refinement pipeline to noisy sources.

\paragraph{Effect of forecast horizon.}
For NSW load and NSW traffic we sweep $H \in \{48, 96, 192, 336, 720\}$ and anneal the modulation caps $\{\lambda_{\max}, \lambda_{\text{ts}}^{\text{cap}}, \lambda_{\text{news}}^{\text{cap}}, g^{\text{cap}}, b^{\text{cap}}\}$ with the horizon (see Appendix~\ref{app:hyperparams}).
The expected pattern is that skill score declines with horizon because the news regime bank's effective causal range is days rather than weeks.


%% ============================================================
%% 7. LIMITATIONS
%% ============================================================
\section{Limitations}
\label{sec:limitations}

\textbf{Heavy-tailed distributions.}
As demonstrated by the NSW Price results, \method struggles with extremely heavy-tailed series where rare spike errors disproportionately impact MAE.
The bounded modulation knobs---while providing safety---limit the delta module's ability to make the large corrections needed during price spikes.

\textbf{Causal chain length.}
The framework is most effective when news has a short causal chain to the target variable (weather $\to$ load, weather $\to$ gas demand).
For variables at the end of longer chains (news $\to$ load $\to$ supply/demand $\to$ price), the regime signal degrades.

\textbf{Daily granularity.}
The regime bank operates at daily granularity, potentially missing intra-day events critical for high-frequency price forecasting.

\textbf{LLM dependency.}
The offline refinement step requires LLM API access.
While this is a one-time cost per corpus and does not affect inference latency, it introduces a dependency.

\textbf{Single-series evaluation.}
Our experiments focus on univariate forecasting.
Extension to multivariate settings remains future work.


%% ============================================================
%% 8. CONCLUSION
%% ============================================================
\section{Conclusion}
\label{sec:conclusion}

We presented \method, a two-stage framework for incorporating LLM-refined news signals into time series forecasting through multiplicative residual modulation.
The key design choice---news signals can only \emph{scale} time-series-derived residual components, never inject additive predictions---provides a formal degradation bound and prevents the model from hallucinating forecasts based on text alone.
Our counterfactual evaluation protocol with news-blanking and date-permutation tests provides causal evidence that improvements are attributable to news content.
Experiments cover five datasets (NSW load, NSW traffic, Netherlands gas demand, Bitcoin, and NSW price), five forecast horizons up to 720 steps, and a mix of curated and open (GDELT) news corpora.
The heavy-tailed price series reveals a \emph{help-rate/MAE decoupling} failure mode that motivates future work on tail-aware modulation strategies and on base architectures that can selectively hand off residual structure to the delta module.

%% ============================================================
%% REFERENCES
%% ============================================================
\bibliographystyle{plainnat}
\bibliography{references}

%% ============================================================
%% APPENDIX
%% ============================================================
\newpage
\appendix

\section{Hyperparameter Details}
\label{app:hyperparams}

Table~\ref{tab:hyperparams} lists the complete hyperparameter settings.
Values reflect the current framework configuration; horizon-specific overrides for Bitcoin and long-horizon sweeps are applied via the run scripts.

\begin{table}[h]
    \centering
    \caption{Complete hyperparameter settings across datasets. Backbone choice is per dataset; all delta modules share the same PatchTST encoder. Bitcoin settings reflect the in-progress run under the rebuilt hourly regime bank.}
    \label{tab:hyperparams}
    \small
    \begin{tabular}{lccccc}
        \toprule
        \textbf{Parameter} & \textbf{Load} & \textbf{Traffic} & \textbf{Gas} & \textbf{Bitcoin} & \textbf{Price} \\
        \midrule
        Base backbone & DLinear & DLinear & DLinear & MLP & DLinear \\
        Base epochs / lr & 40 / $10^{-3}$ & 100 / $10^{-3}$ & 40 / $10^{-3}$ & 40 / $10^{-3}$ & 40 / $10^{-3}$ \\
        Early-stop patience & 5 & 5 & 5 & 5 & 5 \\
        \midrule
        Delta hidden / layers / heads & 128/2/4 & 128/2/4 & 128/2/4 & 128/2/4 & 128/2/4 \\
        Patch len / stride & 8/4 & 8/4 & 8/4 & 8/4 & 8/4 \\
        Delta lr / epochs & $10^{-4}$/30 & $10^{-4}$/50 & $5{\times}10^{-6}$/30 & $10^{-4}$/30 & $5{\times}10^{-6}$/30 \\
        Pretrain epochs / lr & 12/$10^{-3}$ & 12/$10^{-3}$ & 12/$10^{-3}$ & 12/$10^{-3}$ & 12/$10^{-3}$ \\
        Warmup / Grad clip & 10\%/1.0 & 10\%/1.0 & 10\%/1.0 & 10\%/1.0 & 10\%/1.0 \\
        \midrule
        $\lambda_{\min}$/$\lambda_{\text{ts}}^{\text{cap}}$/$\lambda_{\text{news}}^{\text{cap}}$/$\lambda_{\max}$ & .05/.30/.12/.45 & .05/.30/.12/.60 & .05/.30/.12/.45 & .05/.30/.12/.45 & .05/.30/.12/.45 \\
        $g^{\text{cap}}$ / $b^{\text{cap}}$ & 0.20/0.75 & 0.70/0.75 & 0.20/0.75 & 0.20--0.30/0.75 & 0.20--0.30/0.30--0.75 \\
        $\theta$ (active mass) & 0.7 & 0.2 & 0.2 & 0.2 & 0.7 \\
        Spike $k$ / target \% & 3.0/10\% & 3.0/10\% & 3.0/10\% & 3.0/10\% & 4.0/6\% \\
        News blank prob & 30\% & 30\% & 30\% & 30\% & 30\% \\
        Text encoder & E5-sm-v2 & E5-sm-v2 & E5-sm-v2 & E5-sm-v2 & E5-sm-v2 \\
        Regime bank freq. & 30 min & 1 hour & 1 hour & 1 hour & 30 min \\
        News corpus & Curated energy & GDELT 2024 & GDELT 2025 (NL gas) & GDELT 2024 & Curated energy \\
        \bottomrule
    \end{tabular}
\end{table}

\section{Per-Dataset Ablation Tables}
\label{app:ablations}

% [Placeholder: per-dataset tables for backbone, $\theta$, news-corpus, and horizon ablations
% will be populated once the corresponding sweeps are run under the current framework.]

\section{Additional Counterfactual Results}
\label{app:counterfactual}

% [Placeholder: full counterfactual diagnostics (per-dataset, per-horizon) will be added here.
% Headline summary is provided in Table~\ref{tab:counterfactual}.]

\section{NeurIPS Paper Checklist}
\label{app:checklist}

% TODO: Complete NeurIPS paper checklist before submission

\end{document}
